---
title: "16S rRNA Sequence Data Processing"
author: "Brett D. Jameson"
date: "2023-06-05"
output: html_document
---

## Install and load required packages

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
## install packages if you dont already have them in your library
if ("devtools" %in% rownames(installed.packages()) == 'FALSE') install.packages('devtools') 
if ("gridExtra" %in% rownames(installed.packages()) == 'FALSE') install.packages('gridExtra') 
if ("plyr" %in% rownames(installed.packages()) == 'FALSE') install.packages('plyr') 
if ("dplyr" %in% rownames(installed.packages()) == 'FALSE') install.packages('dplyr') 
if ("tidyverse" %in% rownames(installed.packages()) == 'FALSE') install.packages('tidyverse') 

#Read in required libraries

library("devtools")
library("ggplot2")
library("gridExtra")
library('plyr')
library('dplyr')
library("tidyverse")
library("phyloseq")
```

# Saanich Inlet Redox (SaanDox) Experiment: Processing of ASV and taxonomy tables 

This file contains code scripts for cleaning and processing of ASV and taxonomy tables output from QIIME 16S rRNA sequences analysis pipeline. These data were obtained as part of the collaborative research project entitled the Saanich Inlet Redox (SaanDox) Experiment 


## Read Bacteria and Archaea ASV tables, taxonomies, and associated metadata

```{r Read metadata}
metadata.bact <- read.table(file="./data/saandox-seqs-metadata.txt", header = TRUE)
metadata.bact <- metadata.bact %>% mutate(sample.id = gsub("-", ".", sample.id)) %>% # select sampling dates for analysis
  subset(month > 8 | month == 1 | month == 2) 
rownames(metadata.bact) <- as.character(unlist(metadata.bact[, 1])) # convert sample IDs to rownames

metadata.arch <- read.table(file="./data/saandox-seqs-metadata-arch.txt", header = TRUE)
metadata.arch <- metadata.arch %>% mutate(sample.id = gsub("-", ".", sample.id)) %>% # select sampling dates for analysis
  subset(month > 8 | month == 1 | month == 2) 
rownames(metadata.arch) <- as.character(unlist(metadata.arch[, 1])) # convert sample IDs to rownames
```


```{r Read data files}
bact.asv <- read.csv(file = "./data/saandox-bact-asv-table.csv") 
names(bact.asv) <- gsub("X", "", names(bact.asv)) # remove X character from column names
rownames(bact.asv) <- as.character(unlist(bact.asv[, 1])) # Transform to data matrices with taxa as row names
bact.asv <- as.data.frame(bact.asv[, -1])
bact.asv <- as.matrix(bact.asv[colnames(bact.asv)[colnames(bact.asv) %in% rownames(metadata.bact)]]) # select columns that match metadata
reorder.idx <- match(rownames(metadata.bact), colnames(bact.asv)) # create index for ASV table reorder
bact.asv  <- bact.asv[, reorder.idx]
bact.asv <- bact.asv[ , colSums(is.na(bact.asv))==0]

arch.asv <- read.csv(file = "./data/saandox-arch-asv-table.csv") 
names(arch.asv) <- gsub("X", "", names(arch.asv)) # remove X character from column names
rownames(arch.asv) <- as.character(unlist(arch.asv[, 1])) # Transform to data matrices with taxa as row names
arch.asv <- as.data.frame(arch.asv[, -1])
arch.asv <- as.matrix(arch.asv[colnames(arch.asv)[colnames(arch.asv) %in% rownames(metadata.arch)]]) # select columns that match metadata
reorder.idx <- match(rownames(metadata.arch), colnames(arch.asv)) # create index for ASV table reorder
arch.asv  <- arch.asv[, reorder.idx]
arch.asv <- arch.asv[ , colSums(is.na(arch.asv))==0]
```

## Read and clean taxonomy tables

```{r Process taxonomy data, echo=FALSE, include=FALSE}
bact.tax <- read.csv(file = "./data/saandox-bact-taxonomy.csv") 
bact.tax <- bact.tax %>% separate(Taxon,c("Domain","Phylum", "Class", "Order", "Family", "Genus"), sep=";") %>% ## This will remove species column which is not needed
  mutate(Domain = gsub('.*__', '', Domain)) %>% mutate(Phylum = gsub('.*__', '', Phylum)) %>%
  mutate(Class = gsub('.*__', '', Class)) %>% mutate(Order = gsub('.*__', '', Order)) %>%
  mutate(Family = gsub('.*__', '', Family)) %>% mutate(Genus = gsub('.*__', '', Genus)) %>%
  rename(ASV = Feature.ID)

rownames(bact.tax) <- as.character(unlist(bact.tax[, 1]))
bact.tax <- as.matrix(bact.tax[, -1])


arch.tax <- read.csv(file = "./data/saandox-arch-taxonomy.csv") 
arch.tax <- arch.tax %>% separate(Taxon,c("Domain","Phylum", "Class", "Order", "Family", "Genus"), sep=";") %>% ## This will remove species column which is not needed
  mutate(Domain = gsub('.*__', '', Domain)) %>% mutate(Phylum = gsub('.*__', '', Phylum)) %>%
  mutate(Class = gsub('.*__', '', Class)) %>% mutate(Order = gsub('.*__', '', Order)) %>%
  mutate(Family = gsub('.*__', '', Family)) %>% mutate(Genus = gsub('.*__', '', Genus)) %>%
  rename(ASV = Feature.ID)

rownames(arch.tax) <- as.character(unlist(arch.tax[, 1]))
arch.tax <- as.matrix(arch.tax[, -1])
```

## Filter rare taxa using Phyloseq

We will combine the ASV tables and corresponding taxonomy tables into a phyloseq object.

```{r Remove rare taxa}
bact.asv.table = otu_table(bact.asv, taxa_are_rows = TRUE) # Create phyloseq objects
bact.tax.table = tax_table(bact.tax)
bact.phyloseq <- phyloseq(bact.asv.table, bact.tax.table)

## Select taxa that occur greater 10 times in at least 10% of samples
bact.phyloseq.filter <- filter_taxa(bact.phyloseq, function(x) sum(x > 10) > (0.10*length(x)), TRUE)
bact.phyloseq.filter

arch.asv.table = otu_table(arch.asv, taxa_are_rows = TRUE) # Create phyloseq objects
arch.tax.table = tax_table(arch.tax)
arch.phyloseq <- phyloseq(arch.asv.table, arch.tax.table)

## Select taxa that occur greater 10 times in at least 10% of samples
arch.phyloseq.filter <- filter_taxa(arch.phyloseq, function(x) sum(x > 10) > (0.10*length(x)), TRUE)
arch.phyloseq.filter
```

## Save filtered ASV tables

```{r Save clean data}
bact.asv.filter = as(otu_table(bact.phyloseq.filter), "matrix")
bact.asv.filter <- tibble::rownames_to_column(as.data.frame(bact.asv.filter), "ASV")
write_csv(bact.asv.filter, file = "./data/saandox-bact-asv-filter.csv")

bact.tax.clean = as(tax_table(bact.phyloseq.filter), "matrix")
bact.tax.clean <- tibble::rownames_to_column(as.data.frame(bact.tax.clean), "ASV")
write_csv(bact.tax.clean, file = "./data/saandox-bact-taxonomy-filter.csv") # Save cleaned taxonomy table

arch.asv.filter = as(otu_table(arch.phyloseq.filter), "matrix")
arch.asv.filter <- tibble::rownames_to_column(as.data.frame(arch.asv.filter), "ASV")
write_csv(arch.asv.filter, file = "./data/saandox-arch-asv-filter.csv")

arch.tax.clean = as(tax_table(arch.phyloseq.filter), "matrix")
arch.tax.clean <- tibble::rownames_to_column(as.data.frame(arch.tax.clean), "ASV")
write_csv(arch.tax.clean, file = "./data/saandox-arch-taxonomy-filter.csv") # Save cleaned taxonomy table
```

